---
title: "Multi-role Workload Allocation"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Multi-role Workload Allocation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Model introduction

Higher-education teaching teams often need to allocate multiple types of 
work across the same pool of staff or graduate students. A common setting involves three roles:

1. Teaching-assistant work (TA), such as tutorials, laboratory sessions, and other forms of direct teaching support.
2. Grading work (GR), such as marking assignments, tests, and examinations.
3. A third role, E, which captures lighter or more intermittent duties such as exam invigilation, consultation support, and routine administrative tasks.

This model is useful when a department must satisfy exact demand for each 
role while also accounting for role-specific preferences, priority rules 
for lighter duties, and protections for selected cohorts. It can also 
incorporate previous workload and balance total workload over a longer 
planning horizon.

Let there be $N_s$ individuals indexed by $i \in \{1,\ldots,N_s\}$ and $N_j$
courses or tasks indexed by $j \in \{1,\ldots,N_j\}$. For each task-role pair,
where $r \in \{\mathrm{TA}, \mathrm{GR}, \mathrm{E}\}$, the required demand
$d_{j,r}$ must be fully assigned.

For each individual-task pair, $P^{\mathrm{TA}}_{i,j}$ and
$P^{\mathrm{GR}}_{i,j}$ denote the user-supplied TA and grading preference
scores. Higher values indicate stronger preferences. Year of study is denoted
by $y_i$(capped within 1-4), and $s_i$ is a user-configurable score that 
guides E allocation. The model also tracks past semester TA and GR workload, denoted by $t_i^{(1)}$ and $g_i^{(1)}$.

TA and GR may protect different cohorts. Let
$y^\ast_{\mathrm{TA}}, y^\ast_{\mathrm{GR}} \in \{1,2,3,4\}$ denote the
protected years for the two roles. Each role's fairness spread excludes its own
protected cohort. The model only support one protected cohort per role, but the two roles 
may protect the same or different cohorts.

The generalized model balances:

1.  annual TA and GR workload fairness,
2.  role-specific TA and grading preferences,
3.  score-guided allocation of lighter E duties, and
4.  separate TA and GR workload protection for selected cohorts.

# Proposed formulation

The formulation in this section is the proposed generalized model. The
executable package examples later in the vignette still use the historical
`assignment = "phd"` interface until the implementation is updated.

## Objective function

\begin{align*}
X_{i,j,r} &\in \mathbb{Z}_{\ge 0} && \text{units of role } r \text{ in course } j \text{ assigned to student } i \\
t_i^{(2)} &= \sum_{j=1}^{N_j} X_{i,j,\mathrm{TA}} && \text{current-semester TA workload} \\
g_i^{(2)} &= \sum_{j=1}^{N_j} X_{i,j,\mathrm{GR}} && \text{current-semester GR workload} \\
e_i^{(2)} &= \sum_{j=1}^{N_j} X_{i,j,\mathrm{E}} && \text{current-semester E workload} \\
T_i &= t_i^{(1)} + t_i^{(2)} && \text{yearly TA workload} \\
G_i &= g_i^{(1)} + g_i^{(2)} && \text{yearly GR workload} \\
w_i^{\mathrm{TA}} &\ge 0 && \text{slack for the protected cohort's TA soft bound} \\
w_i^{\mathrm{GR}} &\ge 0 && \text{slack for the protected cohort's GR soft bound} \\
T_{\max},T_{\min} &\ge 0 && \text{TA workload bounds outside } y^\ast_{\mathrm{TA}} \\
G_{\max},G_{\min} &\ge 0 && \text{GR workload bounds outside } y^\ast_{\mathrm{GR}}
\end{align*}

\[
\begin{aligned}
\min\quad
&\alpha_{\mathrm{TA}}(T_{\max}-T_{\min})
+\alpha_{\mathrm{GR}}(G_{\max}-G_{\min})\\
&-\beta_{\mathrm{TA}}
  \sum_{i=1}^{N_s}\sum_{j=1}^{N_j}
  P^{\mathrm{TA}}_{i,j}X_{i,j,\mathrm{TA}}\\
&-\beta_{\mathrm{GR}}
  \sum_{i=1}^{N_s}\sum_{j=1}^{N_j}
  P^{\mathrm{GR}}_{i,j}X_{i,j,\mathrm{GR}}\\
&-\phi
  \sum_{i=1}^{N_s}\sum_{j=1}^{N_j}
  s_iX_{i,j,\mathrm{E}}\\
&+\rho_{\mathrm{TA}}
  \sum_{i:y_i=y^\ast_{\mathrm{TA}}}w_i^{\mathrm{TA}}
+\rho_{\mathrm{GR}}
  \sum_{i:y_i=y^\ast_{\mathrm{GR}}}w_i^{\mathrm{GR}}.
\end{aligned}
\]

All objective weights are non-negative and user-specified.
$\alpha_{\mathrm{TA}}$ and $\alpha_{\mathrm{GR}}$ control the importance of
annual workload balance for the two main roles.
$\beta_{\mathrm{TA}}$ and $\beta_{\mathrm{GR}}$ reward role-specific preference
satisfaction. The E term rewards allocation to individuals with larger $s_i$
when $\phi > 0$. Finally, $\rho_{\mathrm{TA}}$ and $\rho_{\mathrm{GR}}$
penalize violations of the protected cohorts' soft workload limits.

## Demand satisfaction

For every task and role, assigned units must match demand:

\[
\sum_{i=1}^{N_s}X_{i,j,r}=d_{j,r},
\quad \forall j,\quad
r\in\{\mathrm{TA},\mathrm{GR},\mathrm{E}\}.
\]

## Role-specific workload spread

Annual TA spread is measured outside the TA-protected cohort:

\[
T_{\min}\le T_i\le T_{\max},
\quad \forall i:y_i\ne y^\ast_{\mathrm{TA}}.
\]

Annual GR spread is measured outside the GR-protected cohort:

\[
G_{\min}\le G_i\le G_{\max},
\quad \forall i:y_i\ne y^\ast_{\mathrm{GR}}.
\]

The two protected years may be the same or different.

## Annual workload equality

Let $C$ denote semester workload capacity per individual. The model fixes each
individual's annual workload total at $2C$:

\[
T_i+G_i+e_i^{(2)}=2C,\quad \forall i.
\]

## Protected-cohort soft upper bounds

Current-semester TA workload is softly capped for the TA-protected cohort:

\[
t_i^{(2)}
\le t_{\max}^{(P)}+w_i^{\mathrm{TA}},
\quad \forall i:y_i=y^\ast_{\mathrm{TA}}.
\]

Current-semester grading workload is separately capped for the GR-protected
cohort:

\[
g_i^{(2)}
\le g_{\max}^{(P)}+w_i^{\mathrm{GR}},
\quad \forall i:y_i=y^\ast_{\mathrm{GR}}.
\]

The slack variables preserve feasibility while making excess workload costly
through the corresponding $\rho$ terms.

## Optional current-semester workload bounds

If supplied, role-wide lower and upper bounds are also imposed:

\begin{align}
t_{\min}^{(2)} \le t_i^{(2)} \le t_{\max}^{(2)}, \quad &\forall i, \\
g_{\min}^{(2)} \le g_i^{(2)} \le g_{\max}^{(2)}, \quad &\forall i, \\
e_{\min}^{(2)} \le e_i^{(2)} \le e_{\max}^{(2)}, \quad &\forall i.
\end{align}

If a bound is omitted, its corresponding constraint is not added.

# Current package examples

The following executable examples demonstrate the current package interface.
They are retained so the vignette remains buildable while the generalized
TA/GR formulation is reviewed.

## E-allocation scoring

`extract_phd_info()` accepts a four-value `s` vector ordered by Years 1 to 4.
The default `c(-1, 0, 1, 2)` reproduces the original seniority scoring:

```{r}
default_inputs <- grouper::extract_phd_info(
  student_df = grouper::phd_students_ex001,
  p_mat = grouper::phd_prefmat_ex001,
  d_mat = grouper::phd_demand_ex001,
  e_mode = "none"
)

default_inputs$s
```

Users can provide a different encoding to control the E-allocation objective:

```{r}
custom_inputs <- grouper::extract_phd_info(
  student_df = grouper::phd_students_ex001,
  p_mat = grouper::phd_prefmat_ex001,
  d_mat = grouper::phd_demand_ex001,
  e_mode = "none",
  s = c(0, 1, 3, 6)
)

custom_inputs$s
```

The score vector affects only the E objective. Protection and TA fairness use
`student_df$year`. The preference matrix is also used exactly as supplied, so
users can choose their own numeric scoring during preprocessing rather than
using the example `3/2/1/-99` encoding.

## Protected cohort

`prepare_phd_model()` and the `prepare_model()` wrapper accept
`protected_year`, which must be one value from 1 to 4. It defaults to Year 1.
For example, the following model protects Year 3:

```{r}
year_3_protected_model <- grouper::prepare_model(
  default_inputs,
  assignment = "phd",
  protected_year = 3,
  t_max_y1 = 1
)
```

The selected cohort receives the soft TA-load bound and its associated slack
penalty. Students from all other years enter the TA fairness spread. The
existing `t_max_y1` argument name is retained for backward compatibility, but
its value applies to whichever cohort is selected by `protected_year`.
